---
title: "spending exploration"
output: html_notebook
---

```{r}
library(tidyverse)
library(janitor)
```

```{r}
federal_spending <- read_csv("data/federal spending by county 2020-2024.csv")
```
```{r}
head(federal_spending)
```

```{r}
colnames(federal_spending)
```
```{r}
glimpse(federal_spending)
```

```{r}
arkansas_spending <- federal_spending %>%
  filter(state_nm == "ARKANSAS")

View(arkansas_spending)
```




```{r}
arkansas_spending %>%
  distinct(cty_name) %>%
  arrange(cty_name)
```



total spending by county across all years
```{r}
arkansas_spending_total <- arkansas_spending %>%
  group_by(cty_name) %>%
  summarise(
    total_contracts = sum(contracts, na.rm = TRUE),
    total_grants = sum(grants, na.rm = TRUE),
    total_direct_payments = sum(direct_payments, na.rm = TRUE),
    total_spending = sum(spending_total, na.rm = TRUE)
  ) %>%
  arrange(desc(total_spending))

View(arkansas_spending_total)
```


remove na row
```{r}
arkansas_spending_total <- arkansas_spending_total %>%
  filter(!is.na(cty_name))
```



```{r}
facilities_with_tracts <- read_csv("data/facilities_with_tracts.csv")
```

```{r}
glimpse(facilities_with_tracts)
```

```{r}
view(facilities_with_tracts)
```


adding county names to facilities with tracts
```{r}
facilities_with_tracts <- facilities_with_tracts %>%
  mutate(
    county_fips = substr(census_tract, 1, 5),
    county_name = case_when(
      county_fips == "05069" ~ "JEFFERSON",
      county_fips == "05125" ~ "SALINE",
      county_fips == "05017" ~ "CHICOT",
      county_fips == "05067" ~ "INDEPENDENCE",
      county_fips == "05059" ~ "HOT SPRING",
      county_fips == "05119" ~ "PULASKI",
      county_fips == "05065" ~ "IZARD",
      county_fips == "05079" ~ "LINCOLN",
      county_fips == "05077" ~ "LEE",
      TRUE ~ "UNKNOWN"
    )
  )
```



```{r}
facilities_with_tracts %>%
  distinct(county_fips, county_name) %>%
  arrange(county_fips)
```


```{r}
# Get distinct prison counties
prison_counties <- facilities_with_tracts %>%
  distinct(county_name) %>%
  pull(county_name)

# Add a flag to your spending table
arkansas_spending_total <- arkansas_spending_total %>%
  mutate(has_prison = ifelse(cty_name %in% prison_counties, "Yes", "No"))

View(arkansas_spending_total)
```



```{r}
arkansas_spending_total %>%
  filter(has_prison == "Yes") %>%
  arrange(desc(total_spending))
```

 compare average spending between prison and non-prison counties
```{r}
arkansas_spending_total %>%
  group_by(has_prison) %>%
  summarise(
    count = n(),
    avg_total_spending = mean(total_spending, na.rm = TRUE),
    median_total_spending = median(total_spending, na.rm = TRUE),
    avg_contracts = mean(total_contracts, na.rm = TRUE),
    avg_grants = mean(total_grants, na.rm = TRUE),
    avg_direct_payments = mean(total_direct_payments, na.rm = TRUE)
  )
```
non-prison counties: 1.08 Billion 
prison counties: 3.64 Billion 


Is little rock in Pulaski county skewing the results too much just cause it's such a huge county. 
```{r}
arkansas_spending_total %>%
  filter(has_prison == "Yes", cty_name != "PULASKI") %>%
  summarise(
    avg_without_pulaski = mean(total_spending, na.rm = TRUE),
    median_without_pulaski = median(total_spending, na.rm = TRUE)
  )
```
This could mean that prison counties aren't receiving a significantly greater amount of federal funding just for having a prison. Pulaski is just huge and just so happens to have a prison. 
