---
title: "Prison facilities and maps"
output: html_notebook
---

Project: Determine the contrast in the demographics in Arkansas prisons versus their surrounding counties using census data.

Also determine the disparity in wages in the prison Census tracts versus the surrounding counties.

Create maps and charts with the results.


#rsw comment: install tidyverse instead, which has dplyr
#see: https://tidyverse.org/
#grey out this chunk after you load it - use a hash tag #
#install janitor to fix the names of columns
```{r}
# install.packages("tidygeocoder")
# install.packages("dplyr")
install.packages("janitor")
```

#rsw comment: load tidyverse
```{r}
library(tidygeocoder)
# library(dplyr)
library(tidyverse)
library(janitor)
```

#rsw comment - make sure to double check you have the facilities and addreses correctly entered. 
#next time, put the names and addresses in a Google Sheet and import that instead. fewer chances of error
```{r}
facilities <- tibble(
  name = c(
    "Barbara Ester Unit",
    "Benny Magness Unit",
    "Benton Unit",
    "Cummins Unit",
    "Delta Regional Unit",
    "East Arkansas Regional Unit",
    "Grimes Unit",
    "J. Aaron Hawkins Sr. Center",
    "Larry B. Norris Unit",
    "McPherson Unit",
    "Ouachita River Correctional Unit",
    "Pine Bluff Unit",
    "Randall L. Williams Correctional Facility",
    "Tucker Unit",
    "Varner/Varner Supermax Unit",
    "Wrightsville Unit"
  ),
  address = c(
    "7500 Corrections Circle, Pine Bluff, AR 71603",
    "10 Prison Circle, Calico Rock, AR 72519",
    "6701 Highway 67, Benton, AR 72015",
    "2540 Hwy 388, Gould, AR 71643",
    "880 East Gaines, Dermott, AR 71638",
    "326 Lee Road 603, Brickeys, AR 72320",
    "300 Corrections Drive, Newport, AR 72112",
    "Hwy 365, Wrightsville, AR 72183",
    "2501 State Farm Rd, Tucker, AR 72168",
    "302 Corrections Drive, Newport, AR 72112",
    "100 Walco Lane, Malvern, AR 72104",
    "890 Free Line Drive, Pine Bluff, AR 71603",
    "7206 West 7th Avenue, Pine Bluff, AR 71603",
    "2400 State Farm Rd, Tucker, AR 72168",
    "320 State Highway 388, Gould, AR 71643",
    "8400 Highway 386, Wrightsville, AR 72183"
  )
)
```


```{r}
geocoded <- facilities %>%
  geocode(
    address,
    method       = "census",
    lat          = latitude,
    long         = longitude,
    full_results = TRUE
  )
```

#feedback from Claude.ai
#The Census geocoder struggles with rural addresses — which prisons often are. A few likely culprits:

#"Hwy", "Rd", "St Farm Rd" — Census geocoder is picky about abbreviations and non-standard road names -->
# Rural routes and highway addresses like Hwy 365 or Lee Road 603 are often not in the Census TIGER database -->
#Small towns like Brickeys and Calico Rock may have incomplete street data -->
#Try switching to the "osm" (OpenStreetMap) method instead — it handles rural addresses much better: -->
```{r}
geocoded1 <- facilities %>%
  geocode(
    address,
    method       = "osm",
    lat          = latitude,
    long         = longitude,
    full_results = TRUE
  ) |> 
  clean_names()
```

#Manually add lat and long. look up coordinates in Google Maps
# Google: lat long 320 State Highway 388, Gould, AR 71643
# The coordinates for 320 State Highway 388, Gould, AR 71643 (the location of the Arkansas Department of Corrections Varner Unit) are approximately 34.04242° N and 91.60889° W.
```{r}
geocoded2 <- geocoded1 |> 
  mutate(
    latitude  = if_else(name_1 == "Varner/Varner Supermax Unit", 34.04242,  latitude),
    longitude = if_else(name_1 == "Varner/Varner Supermax Unit", -91.60889, longitude)
  )

```

#Explanation of the above chunk 
#if_else is a simple three-part structure: condition, value if TRUE, value if FALSE:

#if_else(name_1 == "Varner/Varner Supermax Unit", 34.04242, latitude)
# if this is true ^^^                            use this^  otherwise keep this^


```{r}
no_match <- geocoded %>% filter(is.na(latitude))
print(no_match$name)
```


```{r}
osm_retry <- no_match %>%
  select(name, address) %>%
  geocode(
    address,
    method       = "osm",
    lat          = latitude,
    long         = longitude,
    full_results = TRUE
  )
```

```{r}
names(no_match)

```


```{r}
nrow(no_match)
print(no_match)
```


```{r}
print(osm_retry %>% select(name...1, latitude, longitude))
```


```{r}
manual <- tibble(
  name = c(
    "Cummins Unit",
    "East Arkansas Regional Unit",
    "Varner/Varner Supermax Unit"
  ),
  address = c(
    "2540 Hwy 388, Gould, AR 71643",
    "326 Lee Road 603, Brickeys, AR 72320",
    "320 State Highway 388, Gould, AR 71643"
  ),
  latitude = c(34.0859182097097, 34.84966405170853, 34.04272538283531),
  longitude = c(-91.69990012402901, -90.60006121824772, -91.60824475772544)
)
```



```{r}
osm_clean <- osm_retry %>%
  filter(!is.na(latitude)) %>%
  select(
    name    = name...1,
    address,
    latitude,
    longitude
  )

```


```{r}
census_clean <- geocoded %>%
  filter(!is.na(latitude)) %>%
  select(
    name,
    address,
    latitude,
    longitude
  )
```


```{r}
all_facilities <- bind_rows(census_clean, osm_clean, manual)

print(all_facilities %>% select(name, latitude, longitude))
```



```{r}
library(httr)
library(jsonlite)

get_census_tract <- function(lat, lon) {
  url <- paste0(
    "https://geocoding.geo.census.gov/geocoder/geographies/coordinates?",
    "x=", lon,
    "&y=", lat,
    "&benchmark=Public_AR_Current",
    "&vintage=Current_Current",
    "&layers=Census%20Tracts",
    "&format=json"
  )
  
  response <- GET(url)
  data <- fromJSON(content(response, as = "text"))
  
  tracts <- data$result$geographies$`Census Tracts`
  
  if (length(tracts) > 0) {
    return(tracts$GEOID[1])
  } else {
    return(NA)
  }
}
```


```{r}
all_facilities$census_tract <- mapply(
  get_census_tract,
  all_facilities$latitude,
  all_facilities$longitude
)
```


```{r}
print(all_facilities %>% select(name, latitude, longitude, census_tract))
```


```{r}
getwd()
```

#write all facilities to your github repository folder so it can be shared
```{r}
#write.csv(all_facilities, "C:/Users/casey/Downloads/facilities_with_tracts.csv", row.names = FALSE)
write.csv(all_facilities, "/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Casey Mann/facilities_with_tracts.csv", row.names = FALSE)
```



#rsw comment - you load these once and then gray them out with a comment: #
```{r}

# install.packages("sf")
# install.packages("ggplot2")
```

```{r}
library(tidycensus)
library(sf)
library(ggplot2)
library(dplyr)
```



```{r}
readRenviron("~/.Renviron")
```

#rsw comment - document your work. what is this loading? B01001_001 is the total population estimate 
```{r}
ar_tracts <- get_acs(
  geography = "tract",
  variables = "B01001_001",
  state     = "AR",
  year      = 2022,
  geometry  = TRUE
)
```

#rsw comment: you need to have all_facilities created here. it is not and the code fails.
#it was probably created in a separate notebook. that's not good enough - it needs to be created here
```{r}
facilities_sf <- all_facilities %>%
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs    = 4326
  )
```


#Fix!
```{r}

all_facilities <- read_csv("/Users/gizmofo/Library/CloudStorage/GoogleDrive-robwells@umd.edu/My Drive/DJNF 2026/Casey Mann/facilities_with_tracts.csv")

facilities_sf <- all_facilities %>%
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs    = 4326
  )

```


```{r}
facilities_sf <- st_transform(facilities_sf, crs = st_crs(ar_tracts))
```


```{r}
ggplot() +
  geom_sf(data = ar_tracts, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = facilities_sf, color = "red", size = 2.5, shape = 16) +
  geom_text(
    data  = all_facilities,
    aes(x = longitude, y = latitude, label = name),
    size  = 2.5,
    vjust = -1
  ) +
  labs(
    title    = "Arkansas State Prison Facilities by Census Tract",
    subtitle = "Each polygon represents one census tract",
    caption  = "Sources: AR DOC, U.S. Census Bureau"
  ) +
  theme_void()
```



```{r}
install.packages("ggrepel", repos = "https://cloud.r-project.org")
```


```{r}
library(ggrepel)
```



```{r}
ggplot() +
  geom_sf(data = ar_tracts, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = facilities_sf, color = "red", size = 2.5, shape = 16) +
  geom_label_repel(
    data        = all_facilities,
    aes(x = longitude, y = latitude, label = name),
    size        = 2.5,
    max.overlaps = 20
  ) +
  labs(
    title    = "Arkansas State Prison Facilities by Census Tract",
    subtitle = "Each polygon represents one census tract",
    caption  = "Sources: AR DOC, U.S. Census Bureau"
  ) +
  theme_void()
```



```{r}
install.packages("leaflet")
install.packages("readxl")
```


```{r}
library(leaflet)
library(readxl)
```

#rsw comment: I can't import this spreadsheet - it's on your hard drive. load to github or dropbox and set the file path from the cloud
```{r}
facility_info <- read_excel("AR_DOC_Facilities.xlsx")
```


```{r}
all_facilities <- all_facilities %>%
  left_join(
    facility_info %>% select(`Facility Name`, Capacity),
    by = c("name" = "Facility Name")
  )
```


```{r}
leaflet(all_facilities) %>%
  addTiles() %>%
  addCircleMarkers(
    lng    = ~longitude,
    lat    = ~latitude,
    radius = 6,
    color  = "red",
    popup  = ~paste0(
      "<b>", name, "</b><br>",
      address, "<br>",
      "Capacity: ", Capacity
    )
  )
```


```{r}
library(tidycensus)
library(sf)
library(ggplot2)
library(dplyr)
library(leaflet)
library(readxl)
library(ggrepel)
```


```{r}
ar_tracts <- get_acs(
  geography = "tract",
  variables = "B01001_001",
  state     = "AR",
  year      = 2022,
  geometry  = TRUE
)
```

```{r}
leaflet() %>%
  addProviderTiles("CartoDB.PositronNoLabels") %>%
  addPolygons(
    data        = ar_tracts,
    fillColor   = "white",
    fillOpacity = 0.5,
    color       = "gray50",
    weight      = 0.5
  ) %>%
  addCircleMarkers(
    data   = all_facilities,
    lng    = ~longitude,
    lat    = ~latitude,
    radius = 6,
    color  = "red",
    popup  = ~paste0(
      "<b>", name, "</b><br>",
      address, "<br>",
      "Capacity: ", Capacity
    )
  )
```

```{r}
ar_tracts <- st_transform(ar_tracts, crs = 4326)
```


```{r}
leaflet() %>%
  fitBounds(-94.6, 33.0, -89.6, 36.5) %>%
  addPolygons(
    data        = ar_tracts,
    fillColor   = "white",
    fillOpacity = 1,
    color       = "gray40",
    weight      = 1
  ) %>%
  addCircleMarkers(
    data   = all_facilities,
    lng    = ~longitude,
    lat    = ~latitude,
    radius = 6,
    color  = "red",
    fillColor = "red",
    fillOpacity = 1,
    popup  = ~paste0(
      "<b>", name, "</b><br>",
      address, "<br>",
      "Capacity: ", Capacity
    )
  )
```


```{r}
leaflet() %>%
  fitBounds(-94.6, 33.0, -89.6, 36.5) %>%
  addPolygons(
    data        = ar_tracts,
    fillColor   = "white",
    fillOpacity = 1,
    color       = "gray40",
    weight      = 1
  ) %>%
  addCircleMarkers(
    data   = all_facilities,
    lng    = ~longitude,
    lat    = ~latitude,
    radius = 6,
    color  = "red",
    fillColor = "red",
    fillOpacity = 1,
    popup  = ~paste0(
      "<b>", name, "</b><br>",
      address, "<br>",
      "Capacity: ", Capacity
    )
  )
```

```{r}
my_map <- leaflet() %>%
  fitBounds(-94.6, 33.0, -89.6, 36.5) %>%
  addPolygons(
    data        = ar_tracts,
    fillColor   = "white",
    fillOpacity = 1,
    color       = "gray20",
    weight      = 1.5
  ) %>%
  addCircleMarkers(
    data        = all_facilities,
    lng         = ~longitude,
    lat         = ~latitude,
    radius      = 6,
    color       = "red",
    fillColor   = "red",
    fillOpacity = 1,
    popup       = ~paste0(
      "<b>", name, "</b><br>",
      address, "<br>",
      "Capacity: ", Capacity
    )
  )

my_map
```


```{r}
my_map <- leaflet() %>%
  fitBounds(-94.6, 33.0, -89.6, 36.5) %>%
  addPolygons(
    data        = ar_tracts,
    fillColor   = "white",
    fillOpacity = 1,
    color       = "black",
    weight      = 2
  ) %>%
  addCircleMarkers(
    data        = all_facilities,
    lng         = ~longitude,
    lat         = ~latitude,
    radius      = 6,
    color       = "red",
    fillColor   = "red",
    fillOpacity = 1,
    popup       = ~paste0(
      "<b>", name, "</b><br>",
      address, "<br>",
      "Capacity: ", Capacity
    )
  )

my_map
```

