---
title: "income_ranking"
output: html_notebook
---

This Notebook is used to explore the median income levels throughout the state of Arkansas by county. I will compare the median income of counties with prisons to ones without prisons. The goal of this investigation is to explore the motivations for a town hosting a prison. 

Hypothesis: Counties with prisons should have higher median incomes

install 
```{r}
#install.packages("tidycensus")
#install.packages("tidyverse")

library(tidycensus)
library(tidyverse)
```

B19013_001 is Median household income 
year = 2023 is because it is the most recent 5-year American Community Survey that is available
```{r}
ar_income <- get_acs(
  geography = "county",
  variables = "B19013_001",  
  state     = "AR",
  year      = 2023,          
  survey    = "acs5"
)

ar_income
```


Removing 'Arkansas' suffix from county names column
Renaming columns 
Sorting highest to lowest
Add a ranking column
```{r}
ar_income_ranked <- ar_income %>%
  mutate(county = str_remove(NAME, ", Arkansas")) %>%
  select(GEOID, county, median_income = estimate, margin_of_error = moe) %>%
  arrange(desc(median_income)) %>%
  mutate(rank = row_number()) %>%
  select(rank, GEOID, county, median_income, margin_of_error)

print(ar_income_ranked, n = 75)
```


Now that we've gotten the median incomes for all counties, we will single out the counties with prisons. 
```{r}
prisons <- read_csv("facilities_with_tracts.csv")
```

The facilities csv is listed in census tracts, but we can get county info from that. A census tract include county indentifiers in the number. The first 5 characters identify the state and county. For example a code like 05069002103 breaks down as 05 (Arkansas), 069 (Jefferson County), and 002103
```{r}
prisons_counties <- prisons %>%
  mutate(GEOID = str_sub(census_tract, 1, 5)) %>%  
  distinct(GEOID) %>%
  mutate(has_prison = "Yes")
```

join to ranked income data to compare
```{r}
ar_income_ranked <- ar_income_ranked %>%
  left_join(prisons_counties, by = "GEOID") %>%
  mutate(has_prison = replace_na(has_prison, "No"))
```


count how many prisons are in each county and pull the ranks of each county with a prison
```{r}
prison_counts <- prisons %>%
  mutate(GEOID = str_sub(census_tract, 1, 5)) %>%
  group_by(GEOID) %>%
  summarise(num_prisons = n())

ar_income_ranked %>%
  inner_join(prison_counts, by = "GEOID") %>%
  arrange(rank) %>%
  select(rank, county, median_income, num_prisons)
```

The ranking of the counties with prisons might indicate that there is no economic benefit or disadvantage that comes with hosting a facility. 


